---
title: "Report for Lift the Ban Coalition \n Produced by: Joe Omatoi"
author: "Joe Omatoi"
date: '2022-11-08'
output: 
  rmdformats::robobook:
    code_folding: hide
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r}
# load pacakges ----
library(tidyverse)
library(skimr)
library(sf)
library(leaflet)


# load data ----

# provided data
basic_dat <- read_csv("data/basic_dat.csv") %>% 
  janitor::clean_names() %>% 
  rename(properties_owned = properties_held_by_taxpayer_match_code) %>% 
  rename(address = property_address) %>% 
  rename(number_of_units = relative_size) %>% 
  select(address, community_area, taxpayer_match_code, lat, long, properties_owned, number_of_units) 

# illinois cook community area data from web
map_dat <- read_sf("https://raw.githubusercontent.com/thisisdaryn/data/master/geo/chicago/Comm_Areas.geojson") %>% 
  rename(community_area = community) %>% 
  mutate(community_area = tools::toTitleCase(tolower(community_area))) 

# combined data
dat <- basic_dat %>% 
  left_join(map_dat, by = "community_area")

```

## Overview

This project was undertaken on behalf of the Lift the Ban Coalition, a non-profit organization that aims to lift the ban on rent control in Illinois in order to increase the affordability of housing in the state. However, opponents of rent control argue that it disproportionately harms small, independent landlords. The objective of this project was to assess the number of such landlords who may be affected by rent control and to propose solutions that balance the need for affordable housing with the interests of small-scale property owners. The report presents data in a clear and concise manner to aid the organization in developing effective legislation.

## Data

Two datasets were combined for this project.

The first dataset, provided by the organization, included 129907 rows of properties in Cook county with 13 columns. The dataset was narrowed down so that each observation (property) had the following 6 key attributes:

-   taxpayer match code (identifies landlords; many landlords have different taxpayer names but can only have one taxpayer match code)

-   number of properties owned by the property's landlord

-   latitude

-   longitude

-   community area

The second dataset was the [IL Cook Community Area data](https://raw.githubusercontent.com/thisisdaryn/data/master/geo/chicago/Comm_Areas.geojson%22), which contained supplementary geographic data needed to visualize property characteristics across Cook county in Illinois.

Both datasets were joined by the common community area column.

## Approach

The task follows the next 2 approaches.

1.  Uncover percentages/count of "big" vs "small" landlords

    The aim of this analysis is to accurately define the criteria for identifying a "small" landlord and assess the claim that such landlords will be adversely impacted by rent control measures. To achieve this, the proportion and number of "small" landlords in the dataset were carefully examined. The number of properties owned (with varying thresholds) is used to define "small" landlords who are identified by the taxpayer match code.


2.  Identify areas where "small" landlord properties are a minority

    Given that the objective is to not only enhance affordability of rent, but also safeguard the interests of "small" landlords, the pertinent question is whether it is possible to identify areas where the number of "small" landlords is particularly low.
    
## Exploratory Analysis

All the variables are complete, meaning there are no missingness issues.

```{r}
# missingness check and summary
basic_dat %>% 
  skim_without_charts()
```

Given that a geospatial analysis will be carried out, a map was generated to ensure that each community area had a sufficient number of properties for analysis. As depicted below, a few community areas contain less than or equal to 100 properties, and hence, warrant close scrutiny during the analysis since the sample size may be insufficient to draw any meaningful conclusions. Loop was among the community areas with less than 100 properties, specifically 78. As a result, the percentage of "small" landlord properties in Loop needs to be approached with caution during the analysis.

```{r, fig.align='center'}
# geographic distribution
dat %>% 
  group_by(community_area) %>% 
  mutate(count = n()) %>% 
  select(count, community_area, geometry) %>% 
  distinct(community_area, .keep_all = TRUE) %>% 
  mutate(count = cut(count, c(0, 200, 1000, Inf))) %>% 
  ggplot(aes(fill = count, geometry = geometry)) + 
  geom_sf() +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Number of properties in each community area",
    fill = "Count"
  )
```

## Analysis

### 1. Uncover percentages/count of "big" vs "small" landlords

The interactive tool below displays the percentages and counts of "big" and "small" landlords based on the number of properties owned, which can be adjusted using the slider. By checking the box, viewers can switch between percentage and count.

According to Lift the Ban Coalition, landlords who own six or fewer properties should be considered "small," whereas the standard number recognized in the discussion is 12. Regardless of the threshold selected, the percentage of "small" landlords remains nearly 100%.

This result is supported by an alternative approach that classifies the two groups based on the number of units owned. By checking the "properties/units" option, viewers can switch to this approach. Even with a threshold as low as 15 units to qualify as a "big" landlord, approximately 98% of landlords are still considered "small".

This finding motivates an investigation on the prevalence of properties owned by "small" landlords in different community areas.

```{r}
shinyApp(
  # define ui
  ui <- fluidPage(
    
    titlePanel(""),

    sidebarLayout(
        sidebarPanel(
            sliderInput("threshold",
                        "Adjust number of properties ",
                        min = 1,
                        max = 300,
                        value = 6),
            checkboxInput("raw", "percentage/number"),
            checkboxInput("switch", "properties/units"),
            ),
            mainPanel(
              plotOutput("graph")
            )
    )
  ),
  # define server
  server <- function(input, output) {
  
  # generate pie chart
  output$graph <- renderPlot({
    
    # properties chart ----
    text <- basic_dat %>% 
      distinct(taxpayer_match_code, .keep_all = T) %>% 
      mutate(
        properties_owned = if_else(properties_owned <= input$threshold, paste('<=', input$threshold), paste('>=', input$threshold))
      ) %>% 
      group_by(properties_owned) %>% 
      count() 
    
    text2 <- text %>% 
      mutate(
        x = 'x',
        n = if_else(properties_owned == paste('<=', input$threshold), sum(text$n), n)
      )
    
    text3 <- text %>% 
      mutate(
        n = paste(round(n/sum(text$n)*100, 0), '%')
      )
    
    properties <- basic_dat %>% 
      filter(!is.na(properties_owned)) %>% 
      distinct(taxpayer_match_code, .keep_all = T) %>% 
      mutate(
        x = "x",
        properties_owned = if_else(properties_owned <=input$threshold, paste('<=', input$threshold), paste('>=', input$threshold))
      ) %>% 
      ggplot() +
      aes(x = x, fill = properties_owned) +
      geom_bar(color = "white", alpha = 0.75) +
      annotate(
        geom = "text",
        x = text2$x,
        y = text2$n - text$n/2,
        label = if(input$raw == T) (text$n) else (text3$n),
        size = 10
      ) +
      coord_polar("y") +
      theme(
        # panel.background = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.grid.major.y = element_blank(),
      ) +
      # theme_void() +
      labs(
        fill = "Properties owned",
        x = NULL,
        y = NULL,
        title = paste("Percentage/number of landlords that own greater/less than or equal to", input$threshold, "properties", sep = " ")
      )

    # units chart ----
    text_units <- basic_dat %>% 
      distinct(taxpayer_match_code, .keep_all = T) %>% 
      mutate(
        number_of_units = if_else(number_of_units <= input$threshold, paste('<=', input$threshold), paste('>=', input$threshold))
      ) %>% 
      group_by(number_of_units) %>% 
      count() 
    
    text2_units <- text_units %>% 
      mutate(
        x = 'x',
        n = if_else(number_of_units == paste('<=', input$threshold), sum(text_units$n), n)
      )
    
    text3_units <- text_units %>% 
      mutate(
        n = paste(round(n/sum(text_units$n)*100, 0), '%')
      )
    
    units <- basic_dat %>% 
      filter(!is.na(number_of_units)) %>% 
      distinct(taxpayer_match_code, .keep_all = T) %>% 
      mutate(
        x = "x",
        number_of_units = if_else(number_of_units <=input$threshold, paste('<=', input$threshold), paste('>=', input$threshold))
      ) %>% 
      ggplot() +
      aes(x = x, fill = number_of_units) +
      geom_bar(color = "white", alpha = 0.75) +
      annotate(
        geom = "text",
        x = text2_units$x,
        y = text2_units$n - text_units$n/2,
        label = if(input$raw == T) (text_units$n) else (text3_units$n),
        size = 10
      ) +
      coord_polar("y") +
      theme(
        # panel.background = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.grid.major.y = element_blank(),
      ) +
      # theme_void() +
      labs(
        fill = "Units owned",
        x = NULL,
        y = NULL,
        title = paste("Percentage/number of landlords that own greater/less than or equal to", input$threshold, "units", sep = " ")
      )
    # output chart ----
    if(input$switch == T) (units) else (properties)
  })
  }
)

```

### 2. Identify areas where "small" landlord properties are a minority

This interactive tool displays the proportion of "small" landlord properties across various community areas, allowing users to adjust the threshold for categorizing landlords using the slider. Clicking on each community area displays its name and corresponding percentage. While the majority of landlords are considered "small" overall, closer examination reveals that several community areas have a minority of "small" landlords. The three community areas with the lowest percentages of "small" landlords were Oakland (30% vs 31%), Near South Side (39% vs 46%), and Loop (49% vs 53%), all located in heavily populated areas of Chicago. This underscores the importance of policies in these areas, as they can have a significant impact on a large number of residents.

```{r}
shinyApp(
  # define ui
  ui <- fluidPage(
    
    titlePanel(""),

    sidebarLayout(
        sidebarPanel(
            sliderInput("threshold",
                        "Adjust number of properties ",
                        min = 1,
                        max = 300,
                        value = 6),
            ),
            mainPanel(leafletOutput("distPlot2"))
    )
  ),
  # define server
  server <- function(input, output) {
  
  # generate map
  output$distPlot2 <- renderLeaflet({
    
    leaflet_dat <- dat %>% 
      group_by(community_area) %>% 
      mutate(
        count = if_else(properties_owned <= input$threshold, 1, 0), 
        percentage = round(mean(count)*100),
      ) %>% 
      select(community_area, percentage, geometry) %>% 
      mutate(label = paste0(community_area, ": ", percentage, "%")) %>% 
      distinct() %>% 
      ungroup() %>% 
      st_as_sf()
  
  bins <- c(0, 25, 50, 75, 100)
  pal <- colorBin("YlOrRd", domain = leaflet_dat$percentage, bins = bins)


  leaflet() %>% 
    setView(lng = -87.69, lat = 41.87, zoom = 10) %>% 
    addProviderTiles(
      "OpenStreetMap",
      group = "OpenStreetMap"
    ) %>% 
    addProviderTiles(
      "CartoDB.Positron",
      group = "CartoDB.Positron"
    ) %>% 
    addPolygons(
      data = leaflet_dat,
      fillColor= ~pal(percentage),
      fillOpacity = 0.2,
      color = "grey",
      weight = 1.5,
      label = leaflet_dat$label
    ) %>%
    addScaleBar("bottomleft") %>% 
    addLegend(
      pal = pal, 
      values = leaflet_dat$percentage, 
      opacity = 0.7, 
      title = "% 'small'",
      position = "topright"
      )

  })
  }
)
```

## Conclusion

Overall, while the opposition's argument that rent control may harm "mom and pop" landlords initially appears convincing, a closer examination of individual community areas reveals a different picture. Although the majority of landlords in Illinois Cook are "mom and pop" landlords, they own less than 50% of properties in highly populated areas such as Oakland, Near South Side, and Loop. This finding highlights the importance of considering community-specific factors when crafting rent control policies.

Further analysis could involve using population measures, expected changes in rent prices with rent control, and other relevant data to quantify the potential financial impact of rent control policies in these specific community areas.